Enabling GPU Virtualization in Cloud Environments

### 2015-10-26 paper.tex
ispell pasado

\abstract{The use of accelerators, such as graphics processing units (GPUs), to reduce the execution time of compute-intensive applications has become popular during the past few years. 
These devices increment the computational power of a node thanks to their parallel architecture.
This trend has led cloud service providers as Amazon or middlewares such as OpenStack to add virtual machines (VMs) including GPUs to their facilities   
instances. 
To fullfil these needs, the guest hosts must be equipped with GPUs which, unfortunately, will be barely utilized if a non GPU-enabled VM is running in the host.
The solution presented in this work is based on GPU virtualization and shareability in order to reach an equilibrium between service supply and the applications' demand of accelerators. 
Concretely, we propose to decouple real GPUs from the nodes by using the virtualization technology rCUDA. 
With this software configuration, GPUs can be accessed from any VM avoiding the need  
of placing a physical GPUs in each guest host. 
Moreover, we study the viability of this approach using a public cloud 
service configuration, and we develop a module for OpenStack in order to 
add support for the virtualized devices and the logic to manage them. 
The results demonstrate this is a viable configuration which adds flexibility to current and well-known 
cloud solutions. 
}

@@

\section{\uppercase{Introduction}}
\label{sec:introduction}
Nowadays, many cloud vendors have started offering virtual machines (VMs) with graphics processing units (GPUs) in order to provide
GPGPU (general-purpose GPU) computation services. A few relevant examples include 
Amazon Web Services (AWS)\footnote{\url{https://aws.amazon.com}}, 
Penguin Computing\footnote{\url{http://www.penguincomputing.com}}, 
Softlayer\footnote{\url{http://www.softlayer.com}} and 
Microsoft Azure\footnote{\url{https://azure.microsoft.com}}.
In the public scope, one of the most popular cloud vendors is AWS, which offers a wide range of preconfigured instances ready to be launched.
Alternatively, owning the propper infrastructure, a private cloud can be deployed using a specific middleware such as
Openstack\footnote{\url{https://www.openstack.org}} or Opennebula\footnote{\url{http://opennebula.org}}.

@@

Roughtly speaking these virtualization frameworks 
enable cluster configurations with fewer GPUs than nodes.  The goal is that 
GPU-equipped nodes act as GPGPU servers, yielding a GPU-sharing solution that potentially achieves
a higher overall utilization of the accelerators in the system.

@@

\section{\uppercase{State-of-the-Art}}
\label{sec:state}
Our solutions to the defficiencies exposed in the previous section relies on GPU virtualization, sharing resources in order
to attain a fair balance between supply and demand. 
While several efforts with the same goal have been initiated in the past, as exposed next, 
none of them is as ambitious as ours.

@@

\subsubsection{GPUs}
An instance relies on a VM that runs on a real node with its own virtualized components. 
Therefore AWS can leverage a virtualization framework to offer GPU services to all the instances.
Although the {\tt nvidia-smi} command indicates that the GPUs installed are NVIDIA GRID K520, 
we need to verify that these are non-virtualized devices. For this purpose, we execute the NVIDIA SDK {\tt bandwidthtest}. 
As shown in the Table~\ref{table:bwt}, the bandwidth achieved in this test is higher than the network bandwith, revealing that the accelerator is an actual GPU. 
% \begin{table}[htb]
% \renewcommand{\arraystretch}{1.3}
% \caption{NVIDIA SDK {\tt bandwidthtest} execution transferring 32MB using pageable memory using local GPU.}
% \label{table:bwt}
% \tabcolsep=0.09cm
% \begin{center}\begin{tabular}{cccc}
% Name &  Data Movement & Network & Bandwidth \\ \hline \hline
% g2.2xlarge & Host to Device & High& 3,004 MB/s \\ \hline
% g2.2xlarge & Device to Host & High& 2,809 MB/s\\ \hline
% g2.8xlarge & Host to Device & 10 Gb& 2,814 MB/s\\ \hline
% g2.8xlarge & Device to Host & 10 Gb& 3,182 MB/s\\ \hline
% \end{tabular}\end{center}\end{table}

@@

\subsection{Discussion}
The previous experiments reveal that the AWS GPU-instances are not appropriate for HPC
because neither the network nor the accelerators are powerful enough to deliver high performance
when running compute-intensive parallel applications. 
In other words, AWS is more oriented to offer service than to provide performance.
Also, AWS aims to offer a GPGPU service but fails in offering flexibility.
Concretely, AWS forces the user to choose between a basic instance with a GPU and a powerful instance with 4 GPUs.

Table~\ref{table:awsInstances} shows that the resources of the ``g2.8xlarge'' are quadrupled, but so is the cost per hour.
Therefore, in the case of having other necessities (in terms of the instances type), with GPU virtualization technology we could attach an accelerator to any kind of instance.
Furthermore, reducing the budget destinated to cloud services could be possible by customizing the resources of the available instances.
For example, we can work on an instance with 2 GPUs for only \$ 1.3 by launching 2 ``g2.2xlarge'' and using remote GPUs, avoiding to pay the double for features that we are not willing to use in ``g2.8xlarge''.

The last conclusion is related to GPU-shareability. 
As we deduced, AWS reserves GPU-capable physical machines which will be idle until a GPU-instance request is made.
To maintain an expensive accelerator in a standby state is counter-productive. 
It makes more sense to dedicate less devices, accessible from any machine.

\section{\uppercase{GPGPU Shared Service in OpenStack}}
\label{sec:gpgpuOS}
In this section we review the work related with OpenStack, from the development of an external module to handle remote GPUs to the physical setup and its testing.

\subsection{Managing remote GPUs with OpenStack}
The main idea is to evolve from the original OpenStack architecture (see Figure~\ref{subfig:os-orig})
 to a solution where a new shared service, responsible for all about GPU accelerators, becomes integrated into the architecture (see Figure~\ref{subfig:os-mod}).
This new service brings more flexibility when managing GPUs and new working modes for GPGPU computation in the Cloud.
As illustrated in Figure~\ref{fig:internal}, we alter the original OpenStack Dashboard with a new parser, 
which splits the HTTP query in order to make use of both the GPGPU API for GPU-related operations, and the Nova API for the rest of the computations. 
The new GPGPU Service allows to access GPUs in a VM, but also to create ``GPU-pools'', a set of independent GPUs (logically disattached from the nodes) that can be assigned to one or more VMs.

@@

Thanks to the modular approach of the GPGPU Service, the Nova Project does not need to be modified and the tool can be easily ported to other Cloud Computing solutions.

\subsection{Working Modes}
The developed module allows users to create any of the scenarios displayed in Figure~\ref{fig2}. 
The users are given two configuration options to decide whether a physical GPU will be completely reserved for an instance (first column) or the instance will address a partition of the GPU as if it were a real device (second column).
We refer to this as the ``mode'', with possible values being: ``exclusive'' or ``shared''. 
Let us assume there are 4 GPU devices in the cluster (independently where they were hosted. 
An example of those scenarios is shown in Figure~\ref{fig2}.
There, in ``exclusive'' mode the instance monopolizes all the GPUs, while in the ``shared'' mode, the GPUs are partitioned. 
As a result of halving the GPU memory of the accelerators, the instance will be able to work with up to 8 GPUs, provided that each partition can be addressed as an independent GPU.

Moreover, the users are also responsible for deciding whether a GPU (or a pool) will be assigned to other instances. 
This behaviour is refereed as ``scope'' and it determines that a group of instances is logically connected to a pool of GPUs.
Working with ``public scope'' (bottom row of Figure~\ref{fig2}) implies that the GPUs of a pool can be used simultaneously by all the instances linked to it.
Again, the GPU pool can be composed of ``exclusive'' or ``shared'' GPUs.

@@

\subsection{Discussion}
The network interconnect restricts the performance of our executions.
This is not a new topic and in~\cite{tonithesis} is explained that improving the network infrastructure could make a big different for GPU virtualization.

The most remarkable achievement is the wide range of possible configurations and the flexibility to set and reset a system to fit the 
user requirements.

With this virtualization technology, the requests for GPU devices can be fulfilled with small investments in infrastructure and maintenance.
Energy can be saved not only thanks to the remote access and the ability to emulate several GPUs from a real one, but also  by grouping the accelerators in a single machine (when possible) or turning down nodes when their GPUs are idle.

\section{\uppercase{Conclusions}}
\label{sec:conclusions}
We have presented a complete study of the possibilities offered by AWS when it comes to GPUs. 
The constraints imposed by this service motivated us to deploy our own private cloud, in order to gain flexibility when working with these accelerators.
For this purpose, we have introduced an extension of OpenStack which easily permits to create GPU-instances and manage the physical GPUs to better profit from them.

As we expected, due to the limited bandwidth of the interconnects used in the experimentation, the performance observed for the GPU virtualized scenarios in the test was quite low.
On the other hand we have created new operation modes that open interesting new ways of using GPUs in situations where the having access to a GPU is more important than having a powerful GPU to boost the performance.

\section{\uppercase{Future work}}
\label{sec:future}
The first in the list of pending work is the upgrade of the interconnection network to one more prone on HPC.
Porting the setup and the tests to an infrastructure with an Infiniband network will shed light on the viability of this kind of solutions for HPC.
The same thought, motivates us to try other Cloud vendors which better support for HPC.

Looking for real situations where performance is less important than flexibility will drive us 
to tools for easily deploying GPU-programming computer labs.

Finally, an interesting future mode is to design new strategies in order to decide where a remote GPUs is created and assigned to a physical device
In particular, innovate scheduling policies can enhance the flexibility given by the GPGPU module for OpenStack.

>>>>

\abstract{The use of accelerators, such as graphics processing units (GPUs), to reduce the execution time of compute-intensive applications has become popular during the past few years. 
These devices increment the computational power of a node thanks to their parallel architecture.
This trend has led cloud service providers as Amazon or middlewares such as OpenStack to add virtual machines (VMs) including GPUs to their facilities   
instances. 
To fulfil these needs, the guest hosts must be equipped with GPUs which, unfortunately, will be barely utilized if a non GPU-enabled VM is running in the host.
The solution presented in this work is based on GPU virtualization and shareability in order to reach an equilibrium between service supply and the applications' demand of accelerators. 
Concretely, we propose to decouple real GPUs from the nodes by using the virtualization technology rCUDA. 
With this software configuration, GPUs can be accessed from any VM avoiding the need  
of placing a physical GPUs in each guest host. 
Moreover, we study the viability of this approach using a public cloud 
service configuration, and we develop a module for OpenStack in order to 
add support for the virtualized devices and the logic to manage them. 
The results demonstrate this is a viable configuration which adds flexibility to current and well-known 
cloud solutions. 
}

@@

\section{\uppercase{Introduction}}
\label{sec:introduction}
Nowadays, many cloud vendors have started offering virtual machines (VMs) with graphics processing units (GPUs) in order to provide
GPGPU (general-purpose GPU) computation services. A few relevant examples include 
Amazon Web Services (AWS)\footnote{\url{https://aws.amazon.com}}, 
Penguin Computing\footnote{\url{http://www.penguincomputing.com}}, 
Softlayer\footnote{\url{http://www.softlayer.com}} and 
Microsoft Azure\footnote{\url{https://azure.microsoft.com}}.
In the public scope, one of the most popular cloud vendors is AWS, which offers a wide range of preconfigured instances ready to be launched.
Alternatively, owning the proper infrastructure, a private cloud can be deployed using a specific middleware such as
OpenStack\footnote{\url{https://www.openstack.org}} or Opennebula\footnote{\url{http://opennebula.org}}.

@@

Roughly speaking these virtualization frameworks 
enable cluster configurations with fewer GPUs than nodes.  The goal is that 
GPU-equipped nodes act as GPGPU servers, yielding a GPU-sharing solution that potentially achieves
a higher overall utilization of the accelerators in the system.

@@

\section{\uppercase{State-of-the-Art}}
\label{sec:state}
Our solutions to the deficiencies exposed in the previous section relies on GPU virtualization, sharing resources in order
to attain a fair balance between supply and demand. 
While several efforts with the same goal have been initiated in the past, as exposed next, 
none of them is as ambitious as ours.

@@

\subsubsection{GPUs}
An instance relies on a VM that runs on a real node with its own virtualized components. 
Therefore AWS can leverage a virtualization framework to offer GPU services to all the instances.
Although the {\tt nvidia-smi} command indicates that the GPUs installed are NVIDIA GRID K520, 
we need to verify that these are non-virtualized devices. For this purpose, we execute the NVIDIA SDK {\tt bandwidthtest}. 
As shown in the Table~\ref{table:bwt}, the bandwidth achieved in this test is higher than the network bandwidth, revealing that the accelerator is an actual GPU. 
% \begin{table}[htb]
% \renewcommand{\arraystretch}{1.3}
% \caption{NVIDIA SDK {\tt bandwidthtest} execution transferring 32MB using pageable memory using local GPU.}
% \label{table:bwt}
% \tabcolsep=0.09cm
% \begin{center}\begin{tabular}{cccc}
% Name &  Data Movement & Network & Bandwidth \\ \hline \hline
% g2.2xlarge & Host to Device & High& 3,004 MB/s \\ \hline
% g2.2xlarge & Device to Host & High& 2,809 MB/s\\ \hline
% g2.8xlarge & Host to Device & 10 Gb& 2,814 MB/s\\ \hline
% g2.8xlarge & Device to Host & 10 Gb& 3,182 MB/s\\ \hline
% \end{tabular}\end{center}\end{table}

@@

\subsection{Discussion}
The previous experiments reveal that the AWS GPU-instances are not appropriate for HPC
because neither the network nor the accelerators are powerful enough to deliver high performance
when running compute-intensive parallel applications. 
In other words, AWS is more oriented toward offering a general-purpose service than to provide high performance.
Also, AWS fails in offering flexibility as it forces the user to choose between a basic instance with a GPU and a powerful instance with 4 GPUs.

Table~\ref{table:awsInstances} shows that the resources of the ``g2.8xlarge'' are quadrupled, but so is the cost per hour.
Therefore, in the case of having other necessities (in terms of the instances type), using GPU virtualization technology we could in principle attach an accelerator to any type of instance.
Furthermore, reducing the budget spent in cloud services is possible by customizing the resources of the available instances.
For example, we can work on an instance with 2 GPUs for only \$ 1.3 by launching 2 ``g2.2xlarge'' and using remote GPUs, avoiding to pay the double for features that we do not need in ``g2.8xlarge''.

The last conclusion is related to GPU-shareability. 
As we deduced, AWS reserves GPU-capable physical machines which will be idle until a GPU-instance request is made.
Maintaining an expensive accelerator in a standby state is counter-productive. 
It makes more sense to dedicate less devices, accessible from any machine.

\section{\uppercase{GPGPU Shared Service in OpenStack}}
\label{sec:gpgpuOS}
In this section we describe our work on OpenStack, from the development of an external module to handle remote GPUs to the physical setup and its validation.

\subsection{Managing remote GPUs with OpenStack}
The main idea is to evolve from the original OpenStack architecture (see Figure~\ref{subfig:os-orig})
 to a solution where a new shared service, responsible for the GPU accelerators, becomes integrated into the architecture (see Figure~\ref{subfig:os-mod}).
This new service brings more flexibility when managing GPUs and new working modes for GPGPU computation in the Cloud.
As illustrated in Figure~\ref{fig:internal}, we alter the original OpenStack Dashboard with a new parser, 
which splits the HTTP query in order to make use of both the GPGPU API for GPU-related operations, and the Nova API for the rest of the computations. 
The new GPGPU Service grants access to GPUs in a VM, but also allows the creation of ``GPU-pools'', consisting of a set of independent GPUs (logically disattached from the nodes) that can be assigned to one or more VMs.

@@

Thanks to the modular approach of the GPGPU Service, the Nova Project does not need to be modified, and the tool can be easily ported to other Cloud Computing solutions.

\subsection{Working Modes}
The developed module allows users to set up any of the scenarios displayed in Figure~\ref{fig2}. 
The users are given two configuration options to decide whether a physical GPU will be completely reserved for an instance (first column) or the instance will address a partition of the GPU as if it were a real device (second column).
We refer to this as the ``mode'', with possible values being: ``exclusive'' or ``shared''. 
Let us assume there are 4 GPU devices in the cluster (independently of where they were hosted). 
An example of these scenarios is shown in Figure~\ref{fig2}.
There, in the ``exclusive'' mode the instance monopolizes all the GPUs; while in the ``shared'' mode, the GPUs are partitioned. 
As a result of sharing the GPU memory, the instance will be able to work with up to 8 GPUs, provided that each partition can be addressed as an independent GPU.

Moreover, the users are also responsible for deciding whether a GPU (or a pool) will be assigned to other instances. 
This behavior is refereed as ``scope'', and it determines that a group of instances is logically connected to a pool of GPUs.
Working with the ``public'' scope (bottom row of Figure~\ref{fig2}) implies that the GPUs of a pool can be used simultaneously by all the instances linked to it.
Again, the GPU pool can be composed of ``exclusive'' or ``shared'' GPUs.

@@

\subsection{Discussion}
The network interconnect restricts the performance of our executions.
The analysis in~\cite{tonithesis} reveals that improving the network infrastructure can make a big different for GPU virtualization.

The most remarkable achievement is the wide range of possible configurations and the flexibility to adapt a system to fit the user requirements.

In addition, with this virtualization technology, the requests for GPU devices can be fulfilled with small investment in infrastructure and maintenance.
Energy can be saved not only thanks to the remote access and the ability to emulate several GPUs using only a few real ones, but also by consolidating the accelerators in a single machine (when possible), or turning down nodes when their GPUs are idle.

\section{\uppercase{Conclusions}}
\label{sec:conclusions}
We have presented a complete study of the possibilities offered by AWS when it comes to GPUs. 
The constraints imposed by this service motivated us to deploy our own private cloud, in order to gain flexibility when dealing with these accelerators.
For this purpose, we have introduced an extension of OpenStack which can be easily exploited to create GPU-instances as well as manage the physical GPUs to better profit from them.

As we expected, due to the limited bandwidth of the interconnects used in the experimentation, the performances observed for the GPU virtualized scenarios in the tests were quite low.
On the other hand, we have created new operation modes that open interesting new ways to leverage GPUs in situations where having access to a GPU is more important than having a powerful GPU to boost the performance.

\section{\uppercase{Future work}}
\label{sec:future}
The first item in the list of pending work is an upgrade of the network to an interconnect that is more prone to HPC.
In particular, porting the setup and the tests to an infrastructure with an Infiniband network will shed light on the viability of this kind of solutions for HPC.
Similar reasons, motivate us to try other Cloud vendors which better support for HPC.

Looking for real situations where performance is less important than flexibility will drive us to explore alternative tools to easily deploy GPU-programming computer labs.

Finally, an interesting future work is to design new strategies in order to decide where a remote GPUs is created and assigned to a physical device
Concretely, to innovate scheduling policies can enhance the flexibility offered by the GPGPU module for OpenStack.
